Trauma treatment delay

### 2024-01-11 manuscriptlongversion.Rmd

Introduction: Trauma remains a global health challenge, leading to
millions of deaths each year. The biggest impact can be seen in low to
middle income countries, but in high income countries, trauma systems
have been developed to lower the mortality and morbidity associated with
trauma. To further improve on these systems, conferences are held to
better the care patients receive and to identify errors made, called
opportunities for improvement, one of the more common being delayed
treatment. Aim: This retrospective cohort study aimed to determine the
associations between different patient level factors and receiving
delayed treatment. Material and Methods: This study used the database of
Karolinska university hospital to run a logistic regression linking the
outcome, delayed treatment, to variables, observed patient level
factors. Patients treated for trauma at the hospital between 2012 and
2022 were included, with ISS\>9 and/or trauma code activation. Exclusion
criterias were not being screened for OFI and missing values in any of
the observed variables. Results: Higher ISS and RR was linked to higher
rates of delay in treatment (ISS, OR 1.03, 95% CI 1.01-1.06)(RR, OR
1.04, 95% CI 1.01-1.08), while a lower GCS also was linked to delayed
treatment (OR 0.93, 95% CI 0.87-1.00). Number of injuries showed a
correlation to higher rates of delay, but not when adjusted for. The
remaining variables showed no correlation. Conclusion: Our study
suggests that it is mostly more severely injured and less awake patients
that receive delayed treatment.

@@

@@

The aim was to determine the associations between patient level factors
and receiving delayed treatment in adult trauma patients. A logistic
regression on seven variables showed that higher ISS and RR and lower
GCS was significantly associated with patients receiving delayed
treatment.

@@

Respiratory rate had the correlation that a lower respiratory rate was
associated with lower rates of delayed treatment among the patients. As
the interval is \>10 to \<29 in the triage guideline, both too low and
too high respiratory rate is not desirable[@EHCP2015]. This could have a
similar explanation to ISS and GCS, that patients with tachypnea might
receive higher rates of delayed treatment because they are more severely
injured, therefore a higher risk of missing a possible injury. The
previous study also found this, but the correlation disappeared when
adjusted for, which said study hypothesised being due to its shared
variability with the ISS variable[@Pat_factors]. In this study, the same
correlation could be seen with NOI. A more severely injured patient will
score higher on both ISS and NOI. Therefore, our thought is that the
effect of NOI becomes insignificant when adjusting for ISS.

Not many studies regarding trauma and patient level factors have had
delayed treatment as their outcome, as many are analysing mortality.
Some similarities exist between which patient level factors predict the
outcome, with higher mortality being seen in patients with lower GCS and
higher ISS [@Mort_sammy][@Mort_hefny]. Advanced age has also been
identified as a contributory factor to increased mortality. However, it
is noteworthy that this study did not establish a significant
association between age and delayed treatment. Unlike ISS and GCS, which
are indicative of the severity of injuries and the immediate treatment
needs, advanced age is more likely to introduce comorbidities that
affect overall mortality but may not necessarily heighten the
probability of receiving delayed treatment[@Gioffre2018].

@@

Discussing the generalisability of this study, we included all the
trauma patients treated and Karolinska University Hospital and excluded
40.7% of that population based on that they had not been screened for
OFI. Additional patients were excluded because of missing data, but no
distinct groups except the not screened patients were removed. On that
basis, this population should be a general representation of trauma
patients at the equivalent of a level 1 trauma center, making this study
generalisable. While the population only represents Karolinska
university hospital, the conclusion should be applicable to the whole
trauma population.

@@

I would like to extend my gratitude to my supervisor Martin Gerdin and
my two co-supervisors Jonatan Attergrim and Kelvin Szolnoky for always
being quick with providing valuable help and feedback during all stages
of the writing process as well as guidance during the data analysis part
of the project.

>>>>

Introduction: Trauma remains a global health challenge, leading to
millions of deaths each year. The biggest impact can be seen in low to
middle income countries, but in high income countries, trauma systems
have been developed to lower the mortality and morbidity associated with
trauma. To further improve on these systems, conferences are held to
better the care patients receive and to identify errors made, called
opportunities for improvement, one of the more common being delayed
treatment. Aim: This retrospective cohort study aimed to determine the
associations between different patient level factors and receiving
delayed treatment. Material and Methods: This study used the database of
Karolinska university hospital to run a logistic regression associating the outcome, delayed treatment, to variables, observed patient level
factors. Patients treated for trauma at the hospital between 2012 and
2022 were included, with ISS\>9 and/or trauma code activation. Exclusion
criteria were not being screened for OFI and missing values in any of
the observed variables. Results: Higher ISS (OR 1.03, 95% CI 1.01-1.06)and RR (OR 1.04, 95% CI 1.01-1.08) was linked to higher
rates of delay in treatment while a lower GCS also was linked to delayed
treatment (OR 0.93, 95% CI 0.87-1.00). Number of injuries showed a
correlation to higher rates of delay, but not in the adjusted analysis. The remaining variables were not significantly associated with delayed treatment. Conclusion: Our study suggests that it is mostly more severely injured patients and patients with affected consciousness that receive delayed treatment.

@@

<!-- See edits above -->

@@

The aim was to determine the associations between patient level factors
and receiving delayed treatment in adult trauma patients<!-- Make sure this aim is exactly the same as in the abstract and at the end of the introduction-->. The results indicate that higher ISS and RR and lower
GCS was significantly associated with patients receiving delayed
treatment.

@@

A lower respiratory rate was associated  with lower rates of delayed treatment among the patients. As the interval is \>10 to \<29 in the triage guideline, both too low and too high respiratory rate is not desirable[@EHCP2015]. This could have a similar explanation to ISS and GCS, that patients with tachypnea might have higher rates of delayed treatment because they are more severely injured, therefore a higher risk of missing a possible injury. The previous study also found this, but the association was no longer significant in the adjusted analysis for, which said study hypothesised being due to its shared variability with the ISS variable[@Pat_factors]. In this study, the same correlation could be seen with NOI. A more severely injured patient will score higher on both ISS and NOI. Therefore, our thought is that the effect of NOI becomes insignificant when adjusting for ISS.

Not many studies regarding trauma and patient level factors have had
delayed treatment as their outcome, as many are analysing mortality.
Some similarities exist between which patient level factors predict the
outcome, with higher mortality being seen in patients with lower GCS and
higher ISS [@Mort_sammy; @Mort_hefny]. Advanced age has also been
identified as a contributory factor to increased mortality. However, it
is noteworthy that this study did not establish a significant
association between age and delayed treatment. Unlike ISS and GCS, which
are indicative of the severity of injuries and the immediate treatment
needs, advanced age is more likely to introduce comorbidities that
affect overall mortality but may not necessarily heighten the
probability of receiving delayed treatment[@Gioffre2018].

@@

Discussing the generalisability of this study, we included all the
trauma patients treated and Karolinska University Hospital and excluded
40.7% of that population based on that they had not been screened for
OFI. Additional patients were excluded because of missing data, but no
distinct groups except the not screened patients were removed. On that
basis, this population should be a general representation of trauma
patients at the equivalent of a level 1 trauma center, making this study
generalisable to similar hospitals in high resource settings. While the population only represents Karolinska university hospital, the conclusion should be applicable to the whole trauma population.<!-- Consider removing this last sentence-->

@@

I would like to extend my gratitude to my supervisor Martin Gerdin Wärnberg and my two co-supervisors Jonatan Attergrim and Kelvin Szolnoky for always being quick with providing valuable help and feedback during all stages of the writing process as well as guidance during the data analysis part of the project.
